<!--- % See http://yihui.name/knitr/demo/child/ for documentation on the parent/child document system of knitr -->


```{r, echo=FALSE}
require(nimble)
``` 

# Sequential Monte Carlo, Particle MCMC, Iterated Filtering, and MCEM {#cha:algos-provided}

The NIMBLE algorithm library is growing and currently includes a suite of sequential Monte Carlo (particle filtering) algorithms, particle MCMC for combining particle filters with MCMC, iterated filtering version 2 and Monte Carlo expectation maximization (MCEM) for maximum likelihood estimation, and k-fold cross-validation. 
 
## Particle filters / sequential Monte Carlo and iterated filtering

### Filtering algorithms

NIMBLE includes algorithms for four different types of sequential Monte Carlo (also known as particle filters), which can be used to sample from the latent states and approximate the log likelihood of a state-space model.  These include the bootstrap filter, the auxiliary particle filter, the Liu-West filter, and the ensemble Kalman filter.  The iterated filtering version 2 (IF2) is a related method for maximum-likelihood estimation.  Each of these is built with the eponymous functions `buildBootstrapFilter`, `buildAuxiliaryFilter`, `buildLiuWestFilter`, `buildEnsembleKF`, and `buildIteratedFilter2`.  Each method requires setup arguments  `model` and `nodes`; the latter should be a character vector specifying latent model nodes.  In addition, each method can be customized using a `control` list argument.  Details on the control options and specifics of the algorithms can be found in the help pages for the functions.
  
Once built, each filter can be run by specifying the number of particles.  Each filter has a modelValues object named `mvEWSamples` that is populated with equally-weighted samples from the posterior distribution of the latent states (and in the case of the Liu-West filter, the posterior distribution of the top level parameters as well) as the filter is run.  The bootstrap, auxiliary, and Liu-West filters, as well as the IF2 method, also have another modelValues object, `mvWSamples`.  This has unequally-weighted samples from the posterior distribution of the latent states, along with weights for each particle.  In addition, the bootstrap and auxiliary particle filters return estimates of the log-likelihood of the given state-space model.
  
 We first create a linear state-space model to use as an example for our particle filter algorithms. 
 
```{r, particle_Filter_Chunk, results = "hide"}
# Building a simple linear state-space model. 
# x is latent space, y is observed data
timeModelCode <- nimbleCode({
  x[1] ~ dnorm(mu_0, 1)
  y[1] ~ dnorm(x[1], 1)
  for(i in 2:t){
    x[i] ~ dnorm(x[i-1] * a + b, 1)
    y[i] ~ dnorm(x[i] * c, 1)
  }
  
  a ~ dunif(0, 1)
  b ~ dnorm(0, 1)
  c ~ dnorm(1,1)
  mu_0 ~ dnorm(0, 1)
})

# simulate some data
t <- 25; mu_0 <- 1
x <- rnorm(1 ,mu_0, 1)
y <- rnorm(1, x, 1)
a <- 0.5; b <- 1; c <- 1
for(i in 2:t){
  x[i] <- rnorm(1, x[i-1] * a + b, 1)
  y[i] <- rnorm(1, x[i] * c, 1)
}

# build the model
rTimeModel <- nimbleModel(timeModelCode, constants = list(t = t), 
                          data <- list(y = y), check = FALSE )

# Set parameter values and compile the model
rTimeModel$a <- 0.5
rTimeModel$b <- 1
rTimeModel$c <- 1
rTimeModel$mu_0 <- 1

cTimeModel <- compileNimble(rTimeModel)
```

#### Bootstrap filter
  
Here is an example of building and running the bootstrap filter.
  
```{r, boot_Filter_Chunk}

# Build bootstrap filter
rBootF <- buildBootstrapFilter(rTimeModel, "x", 
                               control = list(thresh = 0.8, saveAll = TRUE, 
                                              smoothing = FALSE))
# Compile filter   
cBootF <- compileNimble(rBootF,project = rTimeModel)
# Set number of particles
parNum <- 5000
# Run bootstrap filter, which returns estimate of model log-likelihood
bootLLEst <- cBootF$run(parNum)
# The bootstrap filter can also return an estimate of the effective 
# sample size (ESS) at each time point
bootESS <- cBootF$returnESS()
```

#### Auxiliary particle filter

Next, we provide an example of building and running the auxiliary particle filter. Note that a filter cannot be built on a model that already has a filter specialized to it, so we create a new copy of our state space model first.
  
```{r, aux_Filter_Chunk, results = "hide"}

# Copy our state-space model for use with the auxiliary filter
auxTimeModel <- rTimeModel$newModel(replicate = TRUE)
compileNimble(auxTimeModel)
# Build auxiliary filter
rAuxF <- buildAuxiliaryFilter(auxTimeModel, "x", 
                              control = list(thresh = 0.5, saveAll = TRUE))
# Compile filter   
cAuxF <- compileNimble(rAuxF,project = auxTimeModel)
# Run auxiliary filter, which returns estimate of model log-likelihood
auxLLEst <- cAuxF$run(parNum)
# The auxiliary filter can also return an estimate of the effective 
# sample size (ESS) at each time point
auxESS <- cAuxF$returnESS()
```

#### Liu and West filter

Now we give an example of building and running the Liu and West filter, which can sample from the posterior distribution of top-level parameters as well as latent states. Note that the Liu-West filter ofen performs poorly and is provided primarily for didactic purposes. The Liu and West filter accepts an additional `params` argument, specifying the top-level parameters to be sampled. 

```{r, lw_Filter_Chunk, results = "hide"}

# Copy model
LWTimeModel <- rTimeModel$newModel(replicate = TRUE)
compileNimble(LWTimeModel)
# Build Liu-West filter, also 
# specifying which top level parameters to estimate
rLWF <- buildLiuWestFilter(LWTimeModel, "x", params = c("a", "b", "c"),
                           control = list(saveAll = FALSE))     
# Compile filter   
cLWF <- compileNimble(rLWF,project = LWTimeModel)
# Run Liu-West filter
cLWF$run(parNum)
```

#### Ensemble Kalman filter

Next we give an example of building and running the ensemble Kalman filter, which can sample from the posterior distribution of latent states. 

```{r, ENKF_Filter_Chunk, results = "hide"}
# Copy model
ENKFTimeModel <- rTimeModel$newModel(replicate = TRUE)
compileNimble(ENKFTimeModel)
# Build and compile ensemble Kalman filter
rENKF <- buildEnsembleKF(ENKFTimeModel, "x",
                         control = list(saveAll = FALSE))  
cENKF <- compileNimble(rENKF,project = ENKFTimeModel)
# Run ensemble Kalman filter
cENKF$run(parNum)
```
  
  Once each filter has been run, we can extract samples from the posterior distribution of our latent states as follows:
  
```{r, particle_Filter_Samples, eval=FALSE}
# Equally-weighted samples (available from all filters)
bootEWSamp <- as.matrix(cBootF$mvEWSamples) # alternative: as.list
auxEWSamp <- as.matrix(cAuxF$mvEWSamples)
LWFEWSamp <- as.matrix(cLWF$mvEWSamples)
ENKFEWSamp <- as.matrix(cENKF$mvEWSamples)

# Unequally-weighted samples, along with weights (available 
# from bootstrap, auxiliary, and Liu and West filters)
bootWSamp <- as.matrix(cBootF$mvWSamples, "x")
bootWts <- as.matrix(cBootF$mvWSamples, "wts")
auxWSamp <-  as.matrix(xAuxF$mvWSamples, "x")
auxWts <- as.matrix(cAuxF$mvWSamples, "wts")

# Liu and West filter also returns samples 
# from posterior distribution of top-level parameters:
aEWSamp <- as.matrix(cLWF$mvEWSamples, "a")
```

#### Iterated filtering 2 (IF2)

The IF2 method accomplishes maximum likelihood estimation using a scheme wherein both latent states and parameters are represented by particles that are weighted and resampled during the iterations.  Iterations include perturbations to the parameter particles following a schedule of decreasing magnitude to yield convergence to the MLE.

Here we apply IF2 to Nile River flow data, specifying a changepoint in the year the Aswan Dam was constructed, as the dam altered river flows.

```{r, if2-model}
library(FKF)

flowCode <- nimbleCode({
    for(t in 1:n)
        y[t] ~ dnorm(x[t], sd = sigmaMeasurements)
    x[1] ~ dnorm(x0, sd = sigmaInnovations)    
    for(t in 2:n)
        x[t] ~ dnorm((t-1==28)*meanShift1899 + x[t-1], sd = sigmaInnovations)
    logSigmaInnovations ~ dnorm(0, sd = 100)
    logSigmaMeasurements ~ dnorm(0, sd = 100)
    sigmaInnovations <- exp(logSigmaInnovations)
    sigmaMeasurements <- exp(logSigmaMeasurements)
    x0 ~ dnorm(1120, var = 100)
    meanShift1899 ~ dnorm(0, sd = 100)
})

flowModel <- nimbleModel(flowCode, data = list(y = Nile),
                 constants = list(n = length(Nile)),
                 inits = list(logSigmaInnovations = log(sd(Nile)),
                              logSigmaMeasurements = log(sd(Nile)),
                              meanShift1899 = -100))
```

Note that the prior distributions for the parameters are not used by IF2, except possibly to obtain boundaries of valid parameter values (not the case here).

Now we build the filter, specifying user-controlled standard deviations (in this case the same as the perturbation `sigma` values) for use in generating the initial particles for the parameters via the `control` list.

```{r, if2-filter}
filter <- buildIteratedFilter2(model = flowModel,
                           nodes = 'x',
                           params = c('logSigmaInnovations',
                                      'logSigmaMeasurements',
                                      'meanShift1899'),
                           baselineNode = 'x0',
                           control = list(sigma = c(0.1, 0.1, 5),
                                          initParamSigma = c(0.1, 0.1, 5)))
cFlowModel <- compileNimble(flowModel)
cFilter  <- compileNimble(filter, project = flowModel)
```

We now run the algorithm with 1000 particles for 100 iterations with the schedule parameter equal to 0.2.

In addition to the estimates, we can extract the values of the log-likelihood, the estimates and the standard deviation of the parameter particles as they evolve over the iterations, in order to assess convergence.


```{r, if2-run}
set.seed(1)
est <- cFilter$run(m = 1000, niter = 100, alpha = 0.2)

cFilter$estimates[95:100,] ## Last 5 iterations of parameter values
cFilter$logLik[90:100] ## Last 5 iterations of log likelihood values
```

Comparing to use of the the Kalman Filter from the FKF package, we see the log-likelihood is fairly similar:


```{r, if2-compare}
dtpred <- matrix(0, ncol = length(Nile))
dtpred[28] <- est[3]
ct <- matrix(0)
Zt <- Tt <- matrix(1)

fkfResult <- fkf(HHt = matrix(exp(2*est[1])),
                 GGt = matrix(exp(2*est[2])),
                 yt = rbind(Nile),
                 a0 = 1120,
                 P0 = matrix(100),
                 dt = dtpred, ct = ct, Zt = Zt, Tt = Tt)
fkfResult$logLik
```

    
### Particle MCMC (PMCMC) {#sec:particle-mcmc}

Particle MCMC (PMCMC) is a method that uses MCMC for top-level model
parameters and uses a particle filter to approximate the time-series
likelihood for use in determining MCMC acceptance probabilities
[@Andrieu_Doucet_Holenstein_2010].  NIMBLE implements PMCMC by
providing random-walk Metropolis-Hastings samplers for model
parameters that make use of particle filters in this way. These
samplers can use NIMBLE's bootstrap filter or auxiliary particle
filter, or they can use a user-defined filter. Whichever filter is
specified will be used to obtain estimates of the likelihood of the
state-space model (marginalizing over the latent states), which is
used for calculation of the Metropolis-Hastings acceptance
probability.  The `RW_PF` sampler uses a univariate normal
proposal distribution, and can be used to sample scalar top-level
parameters.  The `RW_PF_block` sampler uses a multivariate normal
proposal distribution and can be used to jointly sample vectors of
top-level parameters.  The PMCMC samplers can be specified with a call
to `addSampler` with `type = "RW_PF"` or `type = "RW_PF_block"`,
a syntax similar to the other MCMC samplers listed
in Section \@ref(sec:samplers-provided).

The `RW_PF` sampler and `RW_PF_block` sampler can be
customized using the `control` list argument to set the adaptive
properties of the sampler and options for the particle filter
algorithm to be used.  In addition, providing
`pfOptimizeNparticles=TRUE` in the `control` list will use an
experimental algorithm to estimate the optimal number of particles to
use in the particle filter.  See `help(samplers)` for details. The
MCMC configuration for the `timeModel` in the previous section will
serve as an example for the use of our PMCMC sampler.  Here we use the
identity matrix as our proposal covariance matrix.

```{r, pmcmc_Chunk, results = "hide"}
timeConf <- configureMCMC(rTimeModel)   # default MCMC configuration

# Add random walk PMCMC sampler with particle number optimization.
timeConf$addSampler(target = c("a", "b", "c", "mu_0"), type = "RW_PF_block",
                    control = list(propCov= diag(4), adaptScaleOnly = FALSE,
                                 latents = "x", pfOptimizeNparticles = TRUE))
```
  
The  `type = "RW_PF"` and `type = "RW_PF*block"` samplers
default to using a bootstrap filter.  The adapatation control
parameters `adaptive`, `adaptInterval`, and 
`adaptScaleOnly` work in the same way as for an `RW` and `RW*block` samplers.
However, it is not clear if the same approach to adaptation works well
for PMCMC, so one should consider turning off adaptation and using
a well-chosen proposal covariance.

It is also possible that more efficient results
can be obtained by using a custom filtering algorithm.  Choice
of filtering algorithm can be controlled by the `pfType` control
list entry.  The `pfType` entry can be set either to
`'bootstrap'` (the default), `'auxiliary'`, or the name of a
user-defined nimbleFunction that returns a likelihood approximation.

Any user-defined filtering nimbleFunction named in the `pfType`
control list entry must satsify the following:


  1. The nimbleFunction must be the result of a call to
  `nimbleFunction()`.
  1. The `nimbleFunction` must have setup code that accepts the
  following (and only the following) arguments:

      - `model`, the NIMBLE model object that the MCMC algorithm is
  defined on.
      - `latents`, a character vector specifying the latent model
  nodes over which the particle filter will stochastically integrate
  over to estimate the log-likelihood function.
      - `control`, an `R` `list` object.  Note that the
  `control` list can be used to pass in any additional information
  or arguments that the custom filter may require.
  1. The `nimbleFunction` must have a `run` function that
  accepts a single integer arugment (the number of particles to use),
  and returns a scalar double (the log-likelihood estimate).
  1. The `nimbleFunction` must define, in setup code, a
  `modelValues` object named `mvEWSamples` that is used to
  contain equally weighted samples of the latent states (that is, the
  `latents` argument to the setup function).  Each time the
  `run()` method of the `nimbleFunction` is called with number
  of particles `m`, the `mvEWSamples` `modelValues` object
  should be resized to be of size `m` via a call to
  `resize(mvEWSamples, m)`.



## Monte Carlo Expectation Maximization (MCEM)

   Suppose we have a model with missing data (or a layer of latent
  variables that can be treated as missing data), and we would like to
  maximize the marginal likelihood of the model, integrating over the
  missing data. A brute-force method for doing this is MCEM. This is
  an EM algorithm in which the missing data are simulated via Monte
  Carlo (often MCMC, when the full conditional distributions cannot be
  directly sampled from) at each iteration.  MCEM can be slow, and
  there are other methods for maximizing marginal likelihoods that can
  be implemented in NIMBLE.  The reason we started with MCEM is to
  explore the flexibility of NIMBLE and illustrate the ability to combine
  R and NIMBLE to run an algorithm, with R managing the highest-level processing
  of the algorithm and calling nimbleFunctions for computations.  
  
  NIMBLE provides an ascent-based MCEM algorithm, created using `buildMCEM`, that automatically determines when the algorithm has converged by examining 
  the size of the changes in the likelihood between each iteration.  Additionally, the MCEM algorithm can provide an estimate of the asymptotic covariance matrix of the parameters.  An example of calculating the asymptotic covariance can be found in Section \@ref(sec:estimate-mcem-cov).
  
  We will revisit the *pump* example to illustrate the use of
  NIMBLE's MCEM algorithm.
<!--- % newPump didn't exist so I'm creating it in some non-echoed code. -->
```{r, echo=FALSE}
pumpCode <- nimbleCode({ 
  for (i in 1:N){
      theta[i] ~ dgamma(alpha,beta)
      lambda[i] <- theta[i]*t[i]
      x[i] ~ dpois(lambda[i])
  }
  alpha ~ dexp(1.0)
  beta ~ dgamma(0.1,1.0)
})

pumpConsts <- list(N = 10,
                   t = c(94.3, 15.7, 62.9, 126, 5.24,
                       31.4, 1.05, 1.05, 2.1, 10.5))

pumpData <- list(x = c(5, 1, 5, 14, 3, 19, 1, 1, 4, 22))

pumpInits <- list(alpha = 1, beta = 1,
                  theta = rep(0.1, pumpConsts$N))
```   

```{r, build-MCEM, results = "hide", eval = runMCEMs}

pump <- nimbleModel(code = pumpCode, name = "pump", constants = pumpConsts,
                       data = pumpData, inits = pumpInits, check = FALSE)

compileNimble(pump)

# build an MCEM algorithm with ascent-based convergence criterion
pumpMCEM <- buildMCEM(model = pump,
                      latentNodes = "theta", burnIn = 300,
                      mcmcControl = list(adaptInterval = 100),
                      boxConstraints = list( list( c("alpha", "beta"), 
                                                  limits = c(0, Inf) ) ), 
                      buffer = 1e-6)

```

  The first argument `buildMCEM`, `model`, is a NIMBLE model, which can be
  either the uncompiled or compiled version. At the moment, the model provided cannot be part of another MCMC sampler.  

Initial values for the parameters are taken to be the values in the model at the time `buildMCEM` is called, unless the values in the compiled model are changed before running the MCEM.

The ascent-based MCEM algorithm has a number of control options:
 
 
  The `latentNodes` argument should indicate the nodes that will be
  integrated over (sampled via MCMC), rather than
  maximized.  These
  nodes must be stochastic, not deterministic! `latentNodes` will
  be expanded as described in Section \@ref(sec:arbitr-coll-nodes). I.e.,
  either `latentNodes = "x"` or `latentNodes = c("x[1]", "x[2]")`
  will treat `x[1]` and `x[2]` as latent nodes if
  `x` is a vector of two values. All other non-data nodes will be
  maximized over. Note that `latentNodes` can include discrete nodes,
  but the nodes to be maximized cannot.  

 The `burnIn` argument indicates the number of samples from the MCMC for the E-step that should be discarded when computing the expected likelihood in the M-step. Note that `burnIn` can be set to values lower than in standard MCMC computations, as each iteration will start where the last left off. 
  
The  `mcmcControl` argument will be passed to `configureMCMC` to define the MCMC to be used.

 The MCEM algorithm automatically detects box constraints for the nodes that will
 be optimized, using NIMBLE's `getBounds` function.  It is also possible for a user to manually specify constraints via the `boxConstraints` argument.
 Each constraint given should be a list
 in which the first element is the names of the nodes or variables
 that the constraint will be applied to and the second element is a
 vector of length two, in which the first value is the lower limit and
 the second is the upper limit.  Values of `Inf` and `-Inf` are allowed. If a node is not listed, its constraints will be automatically determined by NIMBLE. These constraint arguments are passed as the `lower` and `upper` arguments to R's `optim` function, using `method = "L-BFGS-B"`.  Note that NIMBLE will give a warning if a user-provided constraint is more extreme than the constraint determined by NIMBLE.   

 The value of the  `buffer` argument shrinks the
 `boxConstraints` by this amount.  This can help protect against
 non-finite values occurring when a parameter is on the boundary.
  
  In addition, the MCEM has some extra control options that can be used to further tune the convergence criterion.  See `help(buildMCEM)` for more information.  
  
The `buildMCEM` function returns a list with two elements.  The first element is a function called `run`, which will use the MCEM algorithm to estimate the MLEs.  The second function is called `estimateCov`, and is described in Section \@ref(sec:estimate-mcem-cov).  The `run` function can be run as follows. There is only one run-time argument, `initM`, which is the number of MCMC iterations to use when the algorithm is initialized.

```{r, run-MCEM, eval = runMCEMs}
pumpMLE <- pumpMCEM$run(initM = 1000)
pumpMLE
``` 

Direct maximization after analytically integrating over the latent nodes (possible for this model but often not feasible) gives estimates of $\hat\alpha=0.823$ and $\hat\beta = 1.261$, so the MCEM seems to do pretty well, though tightening the convergence criteria may be warranted in actual usage. 

### Estimating the asymptotic covariance From MCEM {#sec:estimate-mcem-cov}

The second element of the list returned by a call to `buildMCEM` is a function called `estimateCov`, which estimates the asymptotic covariance of the parameters at their MLE values.  If the `run` function has been called previously, the `estimateCov` function will automatically use the MLE values produced by the `run` function to estimate the covariance.  Alternatively, a user can supply their own MLE values using the `MLEs` argument, which allows the covariance to be estimated without having called the `run` function. More details about the `estimateCov` function can be found by calling `help(buildMCEM)`.  Below is an example of using the `estimateCov` function.

```{r, run-MCEM-cov, eval = runMCEMs}
pumpCov <- pumpMCEM$estimateCov()
pumpCov

# Alternatively, you can manually specify the MLE values as a named vector.
pumpCov <- pumpMCEM$estimateCov(MLEs = c(alpha = 0.823, beta = 1.261))
``` 

