% \begin{longtable}[c]{ll}
%  \caption{Basic R manipulation functions in NIMBLE.} \label{table:functions-coreR} \\
%   \hline
  Function & Comments (differences from R) \\
\hline \hline \\
\endhead
\cd{c()} & No \cd{recursive} argument. \\
\cd{rep()} & No \cd{rep.int} or \cd{rep\_len} arguments. \\
\cd{seq()} and `\cd{:}' & Negative integer sequences from `\cd{:}', e.g. \cd{2:1}, do not work. \\
\cd{which()} & No \cd{arr.ind} or \cd{useNames} arguments. \\
\cd{diag()} & Works like R in three ways: \cd{diag(vector)} returns a matrix with \cd{vector}\\
& on the diagonal; \cd{diag(matrix)} returns the diagonal vector of \cd{matrix}; \cd{diag(n)} returns an\\
& $n \times n$ identity matrix.  No \cd{nrow} or \cd{ncol} arguments.\\
\cd{diag()<-} & Works for assigning the diagonal vector of a matrix.\\
\cd{dim()} & Works on a vector as well as higher-dimensional arguments. \\
\cd{length()} & \\
\cd{is.na()} & Does not correctly handle NAs from R that are type \cd{'logical'}, so convert these using \cd{as.numeric()} before passing from R to NIMBLE. \\
\cd{is.nan()} & \\
\cd{numeric()} & Allows additional arguments to control initialization. \\
\cd{logical()} & Allows additional arguments to control initialization. \\
\cd{integer()} & Allows additional arguments to control initialization. \\
\cd{matrix()} & Allows additional arguments to control initialization. \\
\cd{array()} & Allows additional arguments to control initialization. \\
indexing & Arbitrary integer and logical indexing is supported for objects of one or two dimensions. \\
 & For higher-dimensional objects, only `:` indexing works and then only to create an object\\
 &  of at most two dimensions.\\
  \hline \\

%  \end{longtable}
% }